---
title: "Untitled"
author: "Sandra Dela Cruz"
date: "2025-05-06"
output: html_document
---
```{r, include=FALSE}
# Load required packages
library(dplyr)
library(janitor)
library(stringr)
library(ggplot2)
library(tidyr)
library(purrr)
library(DataExplorer)
library(reshape2)
library(pROC) # for ROC/AUC
library(readr)
library(caret) # for confusion matrix
library(DHARMa)
library(lares)
library(mice)
library(gridExtra)
library(nnet) # for multinomial model
library(car) # check for multicollinearity
```

# **Data Exploration**
### **Initial Cleaning Before Splitting**
1. Changed column names to snake format
2. Add number column for row numbers before splitting to ensure no duplicate column is included in the evaluation data
3. Assign class as target variable
4. Rename index to gender
5. Observe data types and detect missing values using summary, no missing values detected
6. Correct variables data types

```{r, echo=FALSE}
library(readr)
dataset_of_diabetes <- read_csv("https://raw.githubusercontent.com/lalaexplore/DATA-621/refs/heads/main/Final%20Project/dataset_of_diabetes.csv")

# 1. Changed column names to snake format
# 2. Add number column for row numbers before splitting
# 3. Assign class as target variable
# "N" (No diabetes)
# "P" (Possibly at risk for diabetes)
# "Y" (Yes, diagnosed with diabetes)
dataset_of_diabetes <- dataset_of_diabetes |>
  mutate(number = row_number()) |>
  clean_names(case = "snake") |>
  rename(target = class, index = 3) |>
  rename(gender = index) |>
  select(id, no_pation, target, everything())

# 5. Observe data types and detect missing values using summary, no missing values detected
summary(dataset_of_diabetes)

# 6. Correct variables data types
dataset_of_diabetes <- dataset_of_diabetes |>
  mutate(
    gender = as.factor(gsub("^f$", "F", gender)),  # Replace lowercase 'f' with 'F'
    target = factor(
      target,
      levels = c("N", "P", "Y")  # No need to make it ordered unless there's a natural order
    )
  )
# Show corrected variables
str(dataset_of_diabetes)

plot_histogram(dataset_of_diabetes)
plot_qq(dataset_of_diabetes)
```

### **Split Dataset into Training and Evaluation**
1. Split original dataset into 8:2
2. Check the distribution of split, output okay
3. Identified overlapping number, none identified

```{r, echo=FALSE}
# using caret package
set.seed(621)
train_target <- createDataPartition(
  dataset_of_diabetes$target, p = 0.8, list = FALSE)
dataset_of_diabetes_training <- dataset_of_diabetes[train_target, ]
dataset_of_diabetes_evaluation <- dataset_of_diabetes[-train_target, ]

# check to ensure the distribution between the split
ggplot(dataset_of_diabetes, aes(x = target, fill = "All")) +
  geom_density(alpha = 0.3) +
  geom_density(data = dataset_of_diabetes_training, aes(
    x = target, fill = "Train"), alpha = 0.5) +
  geom_density(data = dataset_of_diabetes_evaluation, aes(
    x = target, fill = "Evaluation"), alpha = 0.5) +
  labs(title = "Distribution of Target in Train vs Evaluation") +
  theme_minimal()

# Identify overlapping indices
duplicates <- intersect(
  dataset_of_diabetes_training$number, 
  dataset_of_diabetes_evaluation$number)

# Output how many duplicates and optionally which ones
if(length(duplicates) > 0) {
  cat("Found", length(duplicates), "duplicate no_pation:\n")
  print(duplicates)
} else {
  cat("✅ No duplicate indices between training and evaluation sets.\n")
}
```

### **Exploration of Both Datasets**
1. Remove id, no_pation, and number columns because we don't need it
2. Have an overview of both datasets
3. Prior to splitting the datasets into training and evaluation we already performed missing value evaluation and we also corrected the datatypes, and we see that after splitting it did not distort the datatypes.
```{r, echo=FALSE}
# remove id, no_pation, and number columns
dataset_of_diabetes_training <- dataset_of_diabetes_training |> 
  select(-id, -no_pation, -number)
str(dataset_of_diabetes_training)
summary(dataset_of_diabetes_training)


# do the same for evaluation data
dataset_of_diabetes_evaluation <- dataset_of_diabetes_evaluation |>
  select(-id, -no_pation, -number)
str(dataset_of_diabetes_evaluation)
summary(dataset_of_diabetes_evaluation)
```

### **Visual Correlation of Variables**
1. To visualize correlations of variables, we use the `corr_cross()` function from the `lares` package, which allows for an easy comparison. This will potentially help us to include potential features when building our model. We set the method to Spearman since most of our data is not normally distributed (s-shaped as seen in our qq-plot).
```{r, echo=FALSE, message=FALSE}
#| fig-height: 5
#| fig-width: 10

corr_cross(dataset_of_diabetes_training, 
           method = "spearman", max_pvalue = 0.05, top = 20)
```

# **Data Preparation**
1. Drop vldl column because of potential data entry error, and create new column with vldl computed using tg. Vldl = tg/2.2
2. Replace invalid or non-realistic values with NA
Range:
urea = 1-38.9 mg/dL
cr = 26 - 400 umol/L
hb_a1c = 3.7-14%
chol = 3.0-10.3 mmol/L
tg = keep values mmol/L
hdl = 0.4-4.0 mmol/L
ldl = 0.5-8.0 mm/L
vldl = replace with tg/2.2
bmi = keep

```{r, echo=FALSE}
# for training data
dataset_of_diabetes_training <- dataset_of_diabetes_training |>
  select(-vldl) |>  # Drop existing vldl column
  mutate(
    vldl = tg / 2.2,
    urea = ifelse(urea < 1 | urea > 38.9, NA, urea),
    cr = ifelse(cr < 26 | cr > 400, NA, cr),
    hb_a1c = ifelse(hb_a1c < 3.7 | hb_a1c > 14, NA, hb_a1c),
    chol = ifelse(chol < 3.0 | chol > 10.3, NA, chol),
    hdl = ifelse(hdl < 0.4 | hdl > 4.0, NA, hdl),
    ldl = ifelse(ldl < 0.5 | ldl > 8.0, NA, ldl)
    # tg and bmi are kept as-is
  )
summary(dataset_of_diabetes_training)

# for evaluation data
dataset_of_diabetes_evaluation <- dataset_of_diabetes_evaluation |>
  select(-vldl) |>  # Drop existing vldl column
  mutate(
    vldl = tg / 2.2,
    urea = ifelse(urea < 1 | urea > 38.9, NA, urea),
    cr = ifelse(cr < 26 | cr > 400, NA, cr),
    hb_a1c = ifelse(hb_a1c < 3.7 | hb_a1c > 14, NA, hb_a1c),
    chol = ifelse(chol < 3.0 | chol > 10.3, NA, chol),
    hdl = ifelse(hdl < 0.4 | hdl > 4.0, NA, hdl),
    ldl = ifelse(ldl < 0.5 | ldl > 8.0, NA, ldl)
    # tg and bmi are kept as-is
  )
summary(dataset_of_diabetes_evaluation)
```

###**Imputation of Variables**
```{r, echo=FALSE, message=FALSE}
#| fig-height: 5
#| fig-width: 10
# mice imputation
input_training = dataset_of_diabetes_training
method_vector_train <- c("", "", "", "pmm", "pmm",
                   "pmm", "pmm", "", "pmm", "pmm",
                   "", ""
                   )
mice_imp_train = mice(input_training, m = 5, 
                       method = method_vector_train)
summary(input_training$cr)
mice_imp_train$imp$cr
clean_dataset_of_diabetes_training <- complete(mice_imp_train, 3)
summary(clean_dataset_of_diabetes_training)
plot_histogram(clean_dataset_of_diabetes_training)
plot_qq(clean_dataset_of_diabetes_training)

# do the same for evaluation data
input_evaluation = dataset_of_diabetes_evaluation
method_vector_eval <- c("", "", "", "pmm", "pmm",
                   "pmm", "pmm", "", "pmm", "pmm",
                   "", ""
                   )
mice_imp_eval = mice(input_evaluation, m = 5, 
                       method = method_vector_eval)
clean_dataset_of_diabetes_evaluation <- complete(mice_imp_eval, 3)
summary(clean_dataset_of_diabetes_evaluation)
plot_histogram(clean_dataset_of_diabetes_evaluation)
plot_qq(clean_dataset_of_diabetes_evaluation)
```

# **Build Models**
### Model 1
```{r, echo=FALSE, warning=FALSE}

# Fit the multinomial logistic regression model
model1 <- multinom(target ~ gender + hb_a1c + bmi:age + chol, 
                   data = clean_dataset_of_diabetes_training)
cat("Summary Model 1: \n")
summary(model1)

# Get summary of the model (to extract coefficients and standard errors)
summary_model1 <- summary(model1)

# Compute z-scores and p-values
z_scores1 <- summary_model1$coefficients / summary_model1$standard.errors
p_values1 <- 2 * (1 - pnorm(abs(z_scores1)))

# View the p-values
cat("P-values Model 1: \n")
print(p_values1)

library(broom)
tidy(model1)
```

### Model 2
```{r, echo=FALSE, warning=FALSE}

# Fit the multinomial logistic regression model
model2 <- multinom(target ~ gender + hb_a1c + bmi:age + chol:tg, 
                   data = clean_dataset_of_diabetes_training)
cat("Summary Model 2: \n")
summary(model2)

# Get summary of the model (to extract coefficients and standard errors)
summary_model2 <- summary(model2)

# Compute z-scores and p-values
z_scores2 <- summary_model2$coefficients / summary_model2$standard.errors
p_values2 <- 2 * (1 - pnorm(abs(z_scores2)))

# View the p-values
cat("P-values Model 2: \n")
print(p_values2)

library(broom)
tidy(model2)
```
### Model 3
```{r, echo=FALSE, warning=FALSE}

# Fit the multinomial logistic regression model
model3 <- multinom(target ~ gender + hb_a1c + bmi:age + chol:ldl, 
                   data = clean_dataset_of_diabetes_training)
cat("Summary Model 3: \n")
summary(model3)

# Get summary of the model (to extract coefficients and standard errors)
summary_model3 <- summary(model3)

# Compute z-scores and p-values
z_scores3 <- summary_model3$coefficients / summary_model3$standard.errors
p_values3 <- 2 * (1 - pnorm(abs(z_scores3)))

# View the p-values
cat("P-values Model 3: \n")
print(p_values3)

library(broom)
tidy(model3)
```

### Model 4
```{r, echo=FALSE, warning=FALSE}

# Fit the multinomial logistic regression model
model4 <- multinom(target ~ gender + hb_a1c + bmi:age + chol:vldl + cr, 
                   data = clean_dataset_of_diabetes_training)
cat("Summary Model 4: \n")
summary(model4)

# Get summary of the model (to extract coefficients and standard errors)
summary_model4 <- summary(model4)

# Compute z-scores and p-values
z_scores4 <- summary_model4$coefficients / summary_model4$standard.errors
p_values4 <- 2 * (1 - pnorm(abs(z_scores4)))

# View the p-values
cat("P-values Model 4: \n")
print(p_values4)

library(broom)
tidy(model4)
```

# **Select Models**
### Model 1 (Confusion Matrix, ROC curve, and AUC value)
```{r, echo=FALSE}
#| fig-height: 5
#| fig-width: 10
# Predict probabilities on the training data
# Fit the multinomial logistic regression model
model1 <- multinom(target ~ gender + age + hb_a1c + bmi,
                   data = clean_dataset_of_diabetes_training)

# Predict class probabilities
prediction_prob_train1 <- predict(model1, clean_dataset_of_diabetes_training, type = "prob")

# Predict most probable class
prediction_class_train1 <- predict(model1, clean_dataset_of_diabetes_training, type = "class")

# Confusion matrix
conf_matrix_train1 <- confusionMatrix(
  factor(prediction_class_train1, levels = c("N", "P", "Y")),
  factor(clean_dataset_of_diabetes_training$target, levels = c("N", "P", "Y"))
)
print(conf_matrix_train1)

# Accuracy
accuracy1 <- mean(prediction_class_train1 == clean_dataset_of_diabetes_training$target)
cat("Accuracy:", accuracy1, "\n")

# Classification error
classification_error_rate1 <- 1 - accuracy1
cat("Classification Error Rate:", classification_error_rate1, "\n")

# ROC and AUC (One-vs-Rest for each class)
true_labels <- clean_dataset_of_diabetes_training$target
classes <- colnames(prediction_prob_train1)

# Plot setup
par(mfrow = c(1, length(classes)))  # one ROC per class

for (class in classes) {
  # Create binary label: 1 if current class, else 0
  binary_response <- ifelse(true_labels == class, 1, 0)
  pred_probs <- prediction_prob_train1[, class]
  
  roc_curve <- roc(binary_response, pred_probs)
  
  # Plot
  plot(roc_curve,
       main = paste("ROC -", class, "\nAUC:", round(auc(roc_curve), 2)),
       col = "red", lwd = 2)
  
  # Print AUC
  cat("AUC for", class, ":", auc(roc_curve), "\n")
}
```
